# Election Map: 4 Categories

```{r, echo = FALSE, include = FALSE}
fewestVotes <- countyGIS %>% 
  filter(totalVotes == min(totalVotes)) %>%
  select(name, pop, totalVotes)

mostVotes <- countyGIS %>% 
  filter(totalVotes == max(totalVotes)) %>%
  select(name, pop, totalVotes)
```

This map uses 4 custom-labeled factors. 
This map is similar to what you often see on the news, displaying either red or blue. 
However, this one highlights close counties in slightly different colors. 
As you can see, there aren't very many close counties (something that's impossible to see on a two-color map). 




```{r, message=FALSE, warning=FALSE}
pal <- colorNumeric(
  palette = colorRampPalette(c('red', 'white', 'blue'))(length(countyGIS$pctGOP)), 
  domain = countyGIS$pctGOP, reverse=TRUE)

factorPal <- colorFactor(c("blue", "cyan", "magenta", "red"), countyGIS$pctGOPcategories )

leaflet(countyGIS, options = leafletOptions(crsClass = "L.CRS.EPSG3857"), width="100%") %>%
  addPolygons(weight = 0.5, color = "gray", opacity = 0.7,
    fillColor = ~factorPal(pctGOPcategories), fillOpacity = 1, smoothFactor = 0.5,
    label = popupLabels,
    labelOptions = labelOptions(direction = "auto")) %>%
    addPolygons(data = stateGIS,fill = FALSE,color="black",weight = 1) %>%
    addLegend(pal = factorPal,values = ~countyGIS$pctGOPcategories, opacity = 0.7, title = "% Trump",position = "bottomright")

```



What you see is a mostly red country with small pockets of blue and a few close counties. 
What you don't see in this map is anything about population. 
That is a significant limitation to this type of map. 
All counties seem to count equally when you look at the map. 
However, some counties have many times more people than other counties. 
The county with the most votes is `r mostVotes$name` (total population `r format(mostVotes$pop ,big.mark=",")`, total votes `r format(mostVotes$totalVotes,big.mark=",")`).
The county with the fewest votes is `r fewestVotes$name` (total population `r format(fewestVotes$pop ,big.mark=",")`, total votes `r format(fewestVotes$totalVotes,big.mark=",")`).


"Land doesn't vote, people do" is the phrase often associated by the bias created by this type of visualization. 
To get a better sense of this yourself, you can explore the data a bit more. Look at the smallest, say, 1000 counties and see how they voted. Compare that to the largest counties and see how they voted. 
If you start adding up the populations of the smallest counties, how many counties do you have to add up to reach the population of the single largest county? 
Look into a few of these things to get a better sense for the data, and thus for what is conveyed by these types of maps. 
This is not to say that these types of maps aren't useful, but choosing a good color scheme is important, and difficult. 

Here is a specific challenge to test your skills, and get better intuition for how population is distributed in the US. 
Find the population of the largest county.
Figure out how many of the smallest counties you have to add up to reach the population of the largest county (is it the 100 smallest, population-wise, counties, the smallest 500, 1000?).
Then create a map that displays the largest county in one color and all the smallest counties it takes to add up to the same population in a different color. 


